Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex
pass through eta section

\section{Existence of $\beta\eta$-normal forms}
\label{sec:betaeta}
We can easily incorporate the function $\eta$ law to the equational
theory of \lang{} by adding the following parallel reduction rule.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
% \Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
% equational theory since convertibility ($[[a <=> b]]$), the untyped relation
% used for type conversion, is built on top of parallel reduction.
We can recover the same confluence result about parallel reduction
using the standard techniques from
\citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though anti-renaming
(Lemma~\ref{lemma:parantirenaming}) needs to be proven before the
diamond property (Lemma~\ref{lemma:pardiamond}). Another complication
is that confluence and anti-renaming now requires induction on a size
metric of lambda terms so the $\eta$ case has a useful induction hypothesis.

Note that the specification of our logical relation does not require
any changes. We can continue using $\ottkw{ne}$ and $\ottkw{nf}$ to
represent $\beta$-neutral and normal forms. By reproving the
fundamental lemma, we learn that every well-typed term as a
$\beta$-normal form. Lemma~\ref{lemma:parnenf} still holds with our
extended parallel reduction. In the presence of the $\eta$ reduction
rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
preserves $\beta$-normal forms (i.e. does not produce new
$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
functions strictly decreases the size of the term, the existence of
$\beta\eta$ normal form trivially follows.
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}

In \citet{nbeincoq,decagda,martin-lof-a-la-coq}, a relational model is
needed to justify the $\eta$-law for functions. Instead, we prove
confluence for our parallel reduction extended with $\eta$-law for functions
and continue using a logical predicate (i.e. a unary logical relation) to justify
$\eta$-law as part of our equational theory.
A well-known issue with our approach is the failure of syntactic
confluence when the lambda term contains type annotations. A simple
counterexample is $[[\- B ((\- A a) {up 1} var 0)]]$; depending on
whether \rref{P-AbsEta} is performed on the whole term or
\rref{P-AppAbs} is used on the inner $\beta$ redex, we end up with the
terms $[[\- B a]]$ or $[[\- A a]]$, where $[[A]]$ and $[[B]]$ are not
necessarily syntactically equal terms. \citet{choudhury:ddc} resolve
this problem by stating their confluence result in terms of an
equivalence relation that quotients out parts of the terms that are
computationally irrelevant; the annotations of lambda terms are
ignored since the behavior of a lambda term is not affected by its
type annotation. We believe the same approach is applicable to our
proof if we add type annotations to our lambda terms.

The bigger issue is extensions such as $\eta$-laws for unit and
products. Surjective pairing, for example, is an equational law that
is known to cause trouble since it is not confluent for untyped lambda
terms~\citet{KLOP198997}. The relational, type-annotated, and Kripke-style models from
\citet{nbeincoq,decagda,martin-lof-a-la-coq} can be more easily
extended to support these rules.
We note, however, that the issue with $\eta$ rules is not exclusive to dependently
typed languages and has been studied in more limited languages that
are either simply
typed~\citep{pierce2004advanced,pfenning1997computation} or
dependently typed but without large
elimination~\citep{harper2005equivalence,
abel2005untypedconvsurjective}. Common workarounds include
type-directed conversion and shifting the focus to obtaining
$\eta$-long forms~\citet{Abel12}.
We believe our simple proof will help readers
see how techniques from these simpler settings generalize to fully
dependently typed languages and better understand the more complex
mechanized developments.

Finally, due to some technical
details, our mechanized proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & - & - \\
    Renaming & 46 & -  & - \\
    Untyped reduction & 349 & - & - \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \end{minipage}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

>>>>

\section{Existence of $\beta\eta$-normal forms}
\label{sec:betaeta}
\scw{Need an introduction to this section. Why do we care about this new 
rule? What does this section demonstrate?}
\scw{The way we describe this
section should more clearly indicate that we have a mechanized version of this 
proof. And include pointers on where readers might refer to it.}
We can easily incorporate the function $\eta$ law to the equational
theory of \lang{} by adding the following parallel reduction rule.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
% \Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
% equational theory since convertibility ($[[a <=> b]]$), the untyped relation
% used for type conversion, is built on top of parallel reduction.
We can recover the same confluence result about parallel reduction using the
standard techniques from \citet{barendregt:lambda-calculi-with-types,
  takahashi-parallel-reduction}, though anti-renaming
(Lemma~\ref{lemma:parantirenaming}) must be proven before the diamond property
(Lemma~\ref{lemma:pardiamond}). Another complication is that
confluence\scw{diamond property?} and anti-renaming are now proven through
induction on a size metric of lambda terms; \rref{P-AbsEta} reduces a term
that is not a strict subterm.

Note that, after this extension, the specification of our logical relation does not require
any updates. As before, $\ottkw{ne}$ and $\ottkw{nf}$ 
represent $\beta$-neutral and $\beta-$normal forms, and the 
fundamental lemma shows us that every well-typed term has a
$\beta$-normal form. However, in the presence of the $\eta$ reduction
rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
preserves $\beta$-normal forms (i.e. does not produce new
$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
functions strictly decreases the size of the term, the existence of
$\beta\eta$ normal form trivially follows.
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}

%% SCW: This is out of place, and redundant with the next paragraph.
% In contrast to our approach, related work \cite{nbeincoq,decagda,martin-lof-a-la-coq}
% employs a relational model to
% justify the $\eta$-law for functions. 
% Instead, we prove
% confluence for our parallel reduction extended with $\eta$-law for functions
% and continue using a logical predicate (i.e. a unary logical relation) to justify
% $\eta$-law as part of our equational theory.
A well-known issue with our approach is the failure of syntactic
confluence when the lambda term contains type annotations. A simple
counterexample is $[[\- B ((\- A a) {up 1} var 0)]]$; depending on
whether \rref{P-AbsEta} is performed on the whole term or
\rref{P-AppAbs} is used on the inner $\beta$ redex, we end up with the
terms $[[\- B a]]$ or $[[\- A a]]$, where $[[A]]$ and $[[B]]$ are not
necessarily syntactically equal terms. \citet{choudhury:ddc} resolve
this problem by stating their confluence result in terms of an
equivalence relation that quotients out parts of the terms that are
computationally irrelevant; the annotations of lambda terms are
ignored since the behavior of a lambda term is not affected by its
type annotation. We believe the same approach is applicable to our
proof.

The bigger issue is extensions such as $\eta$-laws for unit and
products. Surjective pairing, for example, is not confluent for untyped lambda
terms~\cite{KLOP198997}. The relational, type-annotated, and Kripke-style models from
\citet{nbeincoq,decagda,martin-lof-a-la-coq} can be more easily
extended to support these rules.
We note, however, that the issue with $\eta$ rules is not exclusive to dependently
typed languages and has been studied in more limited languages that
are either simply
typed~\citep{pierce2004advanced,pfenning1997computation} or
dependently typed but without large
eliminations~\citep{harper2005equivalence,
abel2005untypedconvsurjective}. Common workarounds include
type-directed conversion and shifting the focus to obtaining
$\eta$-long forms~\cite{Abel12}.

While not without limitations, our simple proof demonstrates the core
building blocks of more complex arguments, thus paving the way for experimentation 
and eventual extension to more expressive systems.\scw{I tried to reword your 
sentence, but I am still not happy with it.}

\scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this 
sentence elsewhere? I'd like to use the previous sentence as a conclusion for 
this section.}
Finally, due to technical
details, our mechanized proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & - & - \\
    Renaming & 46 & -  & - \\
    Untyped reduction & 349 & - & - \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \end{minipage}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}
